\documentclass{article}

\def\ntitle {Computing Eigensystem}
% \def\nauthor{ } % default author is Zhengbo Zhou
% \def\notes{ } % default is the submitted version
% \def\ndate{ } % default is today's date
\input{clem.tex}

\begin{document}
\maketitle
% \tableofcontents
\thispagestyle{firstpage} 

%%%%%%%%%%%%%%%%%%%%%%%%%%%% MAIN ARTICLE %%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Divide and Conquer}
\label{sec:divide-conquer}
\subsection{Introduction}
This is currently the fastest method to find all the eigenvalues and
eigenvectors of symmetric tridiagonal matrices larger than $n = 25$.
In the worst cases, divide-and-conquer requires $O(n^3)$ flops, but in
practice, the constant is quite small. 
Over a large of set of random test cases, it appears to take only
$O(n^{2.3})$ flops on average, and as low as $O(n^2)$ for some eigenvalue
distribution.

In theory, divide-and-conquer could be implemented to run on $O(n\log^pn)$
flops, where $p$ is a small integer. This super-fast implementation uses
the fast multi-pole method (FMM), originally invented for the completely
different problem of computing the mutual forces on $n$ electrically
charged particles.
But the complexity of this super-fast implementation means that QR
iteration is currently the algorithm of choice for finding all eigenvalues,
and divide-and-conquer without the FMM is the method of choice for finding
all eigenvalues and all eigenvectors.

\subsection{Overview}
It is quite subtle to implement in a numerically stable way. Indeed,
although this method was first introduced in 1981, the ``right''
implementation was not discovered until 1992.  This routine is available as
LAPACK routines \texttt{ssyevd} for dense matrices and \texttt{sstevd} for
tridiagonal matrices.  This routine uses divide-and-conquer for matrices of
dimension larger than 25 and automatically switches to QR iteration for
smaller matrices.

\begin{equation}
  \begin{aligned}
    T & =
        \left[
        \begin{array}{cccc|cccc}
          a_1 & b_1 & & & & & & \\
          b_1 & \ddots & \ddots & & & & & \\
              & \ddots & a_{m-1} & b_{m-1} & & & & \\
              & & b_{m-1} & a_m & b_m & & & \\
          \hline & & & b_m & a_{m+1} & b_{m+1} & & \\
              & & & & b_{m+1} & \ddots & & \\
              & & & & & & \ddots & b_{n-1} \\
              & & & & & & b_{n-1} & a_n
        \end{array}\right]\\
      & =\left[
        \begin{array}{cccc|cccc}
          a_1 & b_1 & & & & & & \\
          b_1 & \ddots & \ddots & & & & \\
              & \ddots & a_{m-1} & b_{m-1} & & & & \\
              & & b_{m-1} & a_m-b_m & & & \\
          \hline & & & & a_{m+1}-b_m & b_{m+1} & & \\
              & & & & b_{m+1} & \ddots & & \\
              & & & & & \ddots & b_{n-1} \\
              & & & & & b_{n-1} & a_n
        \end{array}\right]
        + \left[
        \begin{array}{cc|cc}
          & 	&  &  \\
          & b_m & b_m &  \\\hline
          & b_m & b_m &  \\
          &  & &
        \end{array}\right] \\
      & = \left[\begin{array}{c|c}
                  T_1 & 0 \\
                  \hline 0 & T_2
                \end{array}
        \right]+b_m \cdot\left[
        \begin{array}{c}
          0 \\
          \vdots \\
          0 \\
          1 \\
          1 \\
          0 \\
          \vdots \\
          0
        \end{array}\right][0, \ldots, 0,1,1,0, \ldots, 0] \equiv\left[
        \begin{array}{c|c}
          T_1 & 0 \\
          \hline 0 & T_2
        \end{array}\right]+b_m v v^T .
  \end{aligned}
\end{equation}
Suppose that we have the eigendecomposition of $T_1$ and $T_2$:
$T_i = Q_i\Lambda_iQ_i\tp$. These will be computed recursively by this same
algorithm. We related the eigenvalues of $T$ to those of $T_1$ and $T_2$ as
follows:
\begin{equation}
  \notag
  \begin{aligned}
    T & =
        \begin{bmatrix}
          T_1 & 0 \\ 0 & T_2
        \end{bmatrix}
        + b_mvv\tp \\
      & =
        \begin{bmatrix}
          Q_1\Lambda_1Q_1\tp & 0 \\ 0 & Q_2\Lambda_2Q_2\tp
        \end{bmatrix}
        + b_mvv\tp\\
      & =
        \begin{bmatrix}
          Q_1 & 0 \\ 0 & Q_2
        \end{bmatrix}
        \left(
        \begin{bmatrix}
          \Lambda_1 & \\ & \Lambda_2
        \end{bmatrix} + b_muu\tp
        \right)
        \begin{bmatrix}
          Q_1\tp & 0 \\ 0 & Q_2\tp
        \end{bmatrix}
  \end{aligned}
\end{equation}
where
\begin{equation*}
  u =
  \begin{bmatrix}
    Q_1\tp & 0 \\ 0 & Q_2
  \end{bmatrix}, \qquad
  v =
  \begin{bmatrix}
    \text{last column of $Q_1\tp$} \\ \text{first column of $Q_2\tp$}
  \end{bmatrix}
\end{equation*}
since $v = [0,\dots,0,1,1,0,\dots,0]\tp$. Therefore, the eigenvalues of $T$
are the same as those of the similar matrix $D + \rho uu\tp$ where $D =
\begin{bmatrix}
  \Lambda_1 & 0 \\ 0 & \Lambda_2
\end{bmatrix}
$ is diagonal, $\rho = b_m$ is a scalar, and $u$ is a vector. Henceforth,
we will assume without loss of generality that the diagonal
$d_1,\cdots,d_n$ of $D$ is sorted: $d_n\leq\cdots\leq d_n$.

To find the eigenvalues of $D + \rho uu\tp$, assume first that
$D - \lambda I$ is nonsingular, and compute the characteristic polynomial
as follows:
\begin{equation}
  \label{eq:3}
  \det(D + \rho uu^T - \lambda I) 
  = \det((D - \lambda I)(I + \rho(D - \lambda I )^{-1}uu^T)).
\end{equation}

Since $D - \lambda I$ is nonsingular, $\det(I + \rho(D -
\lambda I)^{-1}uu^T = 0$ whenever $\lambda$ is an eigenvalue. Notice that
$I+\rho(D - \lambda I)^{-1}uu^T$ is the identity plus a rank-1 matrix; the
determinant of such a matrix is easy to compute:

\begin{lemma}
  If $x$ and $y$ are vectors, $\det(I + xy^T) = 1 + y^Tx$.
\end{lemma}
Therefore
\begin{equation}\notag
  \begin{aligned}
    \det(I + \rho(D - \lambda I)^{-1}uu^T)  & = 1 + \rho u^T(D - \lambda
                                              I)^{-1}u \\
      & = 1 + \rho\sum_{i=1}^n \frac{u_i^2}{d_i-\lambda} \equiv f(\lambda). 
  \end{aligned}
\end{equation}
and the eigenvalues of $T$ are the roots of the so-called secular equation
$f(\lambda) = 0$. If all $d_i$ are distinct and all $u_i\neq 0$ (the
generic case), the function $f(\lambda)$ has the graph shown in.



%%%%%%%%%%%%%%%%%%%%%%%%%%% Bibliographies %%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\bibliographystyle{nj-plain}
\bibliography{~/bibtex/references.bib}


\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
